---
title: "Evals"
author: "Roman Tsegelskyi "
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(pander)
```

`evals` is aimed at collecting as much information as possible while evaluating R code. It can evaluate a character vector of R expressions, and it returns a list of captured information while running those:

* `src` holds the R expression,
* `result` contains the raw R object as is,
* `output` represents how the R object is printed to the standard output,
* `type` is the class of the returned R object,
* `msg` is a list of possible messages captured while running the R expression and
* `stdout` contains if anything was written to the standard output.

Besides capturing evaluation information, `evals` is able to automatically identify if an R expression is returning anything to a graphical device, and can save the resulting image in a variety of file formats.

Another interesting feature that `evals` provides is caching the results of evaluated expressions. For more details read the section about Caching.

## Basic Usage

Let's start with basic example of evaluating `1:10` and collecting all information about it:

```{r}
evals('1:10')
```

It is possible to capture only some of the information, by specifying `output` parameter:

```{r}
evals('1:10', output = c('result', 'output'))
```

### Errors/Warings
```{r}
evals('x')[[1]]$msg
```

### Graps and Graphical Options

### General graph capture

### Style unification

## Logging

To make execution and debug more understandable `evals` provides a possibility for logging with `log` parameter. Logging in `evals` relies on [futile.logger](cran.r-project.org/web/packages/futile.logger/futile.logger.pdf) package which provides logging API similar to `log4j`. Basic example:

```{r}
x <- evals('1:10', log = 'foo')
```

`futile.logger`'s thresholds range from most verbose to least verbose: `TRACE`, `DEBUG`, `INFO`, `WARN`, `ERROR`, `FATAL`. By default threshold is set to `INFO` which will hide some unessential information. To permanently set the threshold for logger use `flog.threshold`:

```{r}
evalsOptions('log', 'evals')
flog.threshold(TRACE, 'evals')
x <- evals('1:10', cache.time = 0)
```

`futile.logger` also provides a very useful functionality of write logs to files instead of printing them to promt:

```{r}
t <- tempfile()
flog.appender(appender.file(t), name = 'evals')
x <- evals('1:10', log = 'evals')
readLines(t)
```

````{r, echo = FALSE, message = FALSE}
x <- capture.output(appender.console())
```

## Result Caching

`evals` is using a custom caching algorithm to cache the results of evaluated R expressions.

### How it works

* All R code passed to `evals` is split into single expressions and parsed.
* For each R expression (function call, assignment, etc.), `evals` extracts symbols in a separate list in `getCallParts`. This list describes the unique structure and the content of the passed R expressions
* A hash is computed for each list element and cached too in `pander`'s local environments. This is useful if you are using large data frames, just imagine: the caching algorithm would have to compute the hash for the same data frame each time it's touched! This way the hash is recomputed only if the R object with the given name is changed.
* The list of such R objects is serialized, then an SHA-1 hash is computed taking into consideration `panderOptions` and `evalsOptions`, which all together is unique and there is no real risk of collision.
* If `evals` can find the cached results in the appropriate environment (if `cache.mode set` to enviroment) or in a file named to the computed hash (if `ċache.mode` set to `disk`), then it is returned on the spot. The objects modified/created by the cached code are also updated.
* Otherwise the call is evaluated and the results and the modified R objects of the environment are optionally saved to cache (e.g. if `cache` is active and if the evaluation `proc.time()` > `cache.time` parameter). Cached results are saved in `cached.results` in `pander`'s namespace. `evals` also remembers if R expressions change the evaluation environment (for example assignments) and saves such changes in `cached.environemnts` in `pander`'s namespace.

### Examples

Let's start with small example. We will set `cache.time` to 0, to cache all expressions regardless of time they took to evaluate.

```{r, echo = FALSE, message = FALSE}
evalsOptions('cache.time', 0)
```
```{r}
system.time(evals('1:1e5'))
system.time(evals('1:1e5'))
```

Since the hash for caching computed based on the *structure* and *content* of the R commands instead of the used variable names or R expressions, `evals` is able to achieve great results:

```{r}
x <- mtcars$hp
y <- 1e4
system.time(evals('sapply(rep(x, y), mean)'))
```

Let us create some custom functions and variables, which are not identical to the above call:

```{r}
f <- sapply
g <- rep
h <- mean
X <- mtcars$hp * 1
Y <- 10000
system.time(evals('f(g(X, Y), h)'))
```

Another important feature of `evals` is that it takes notes the changes in the evaluation environment. For example:

```{r}
x <- 1
system.time(evals('x <- 1:10; x[2] - 1'))
```

`x <- x + 2` will be cached, so if the same assignment occurs again we won't need to evaluate it. But what about the change of `x` when we get the result from the cache? `evals` takes care of that.

```{r}
system.time(evals('x <- 1:10; x[3]'))
```

