---
title: "Evals"
author: "Roman Tsegelskyi "
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Evals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(pander)
```

`evals` is aimed at collecting as much information as possible while evaluating R code. It can evaluate a character vector of R expressions, and it returns a list of captured information while running those:

* `src` holds the R expression,
* `result` contains the raw R object as is,
* `output` represents how the R object is printed to the standard output,
* `type` is the class of the returned R object,
* `msg` is a list of possible messages captured while running the R expression and
* `stdout` contains if anything was written to the standard output.

Besides capturing evaluation information, `evals` is able to automatically identify if an R expression is returning anything to a graphical device, and can save the resulting image in a variety of file formats.

Another interesting feature that `evals` provides is caching the results of evaluated expressions. For more details read the section about Caching.

## Basic Usage

Let's start with basic example of evaluating `1:10` and collecting all information about it:
```{r}
evals('1:10')
```

### Errors/Warings
```{r}
evals('x')[[1]]$msg
```

### Graps and Graphical Options

### General graph capture

### Style unification

## Result Caching

### How it works

### Examples
