---
title: "Pandoc.table"
author: "Roman Tsegelskyi "
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pandoc.table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(pander)
```

Core functionality of `pander` is centered around `pandoc.table` which is aimed at rendering tables in `markdown`. `pandoc.table` has big variety of options ((highlighting)[#cell-highlighting], (styles)[#table-styles], etc.) and this vignette aims to give a more details overview of most common options.

## Table styles

Since `pander` aims at rendering R objects into [Pandoc](http://pandoc.org/)'s markdown all four (`multiline`, `simple`, `grid`, `rmarkdown`) of  [Pandoc](http://johnmacfarlane.net/pandoc/README.html#tables)'s formats are supported. General advices is to stick with default `multiline` style, but there is a need to change it either specify `style` argument when calling `pander`/`pandoc.table` or change default style using `panderOptions`.

### multiline tables

`multiline` tables allow headers and table rows to span multiple lines of text (but cells that span multiple columns or rows of the table are not supported). Also note, that, by default, line breaks are removed from cells for simplicity, so in order to have multiline cells `keep.line.breaks` needs to be set to `TRUE`:

```{r}
m <- data.frame('Value\n1', 'Value\n2')
colnames(m) <- c('Multiline\nCol1', 'Multiline\nCol2')
pandoc.table(m, keep.line.breaks = TRUE)
m <- mtcars[1:3, 1:4]
pandoc.table(m)
```

### simple tables

`simple` tables are have more compact syntax that all other styles, but don't they support multiline cells:

```{r, error=TRUE}
m <- mtcars[1:3, 1:4]
pandoc.table(m, style = 'simple')
m <- data.frame('Value\n1', 'Value\n2')
colnames(m) <- c('Multiline\nCol1', 'Multiline\nCol2')
pandoc.table(m, keep.line.breaks = TRUE, style='simple')
```

### grid tables

`grid` format is really handy for `emacs` users ([Emacs table mod](http://table.sourceforge.net/)) and it does support block elements (multiple paragraphs, code blocks, lists, etc.) inside cells. Alignments are not supported by most parsers, meaning that even though `pander` will produce a table with alightnment, it will be lost during conversion from `markdown` to `pdf/html/docx`, and cells can't span multiple columns or rows.

```{r, error=TRUE}
m <- mtcars[1:3, 1:4]
pandoc.table(m, style = 'grid')
m <- data.frame('Value\n1', 'Value\n2')
colnames(m) <- c('Multiline\nCol1', 'Multiline\nCol2')
pandoc.table(m, keep.line.breaks = TRUE, style='grid')
```

### rmarkdown tables

`rmarkdown` or pipe table format is often used directly with `knitr`, since it was supporters by the first versions of the `markdown` package. The beginning and ending pipe characters are optional, but pipes are required between all columns:

```{r, error=TRUE}
m <- mtcars[1:3, 1:4]
pandoc.table(m, style = 'rmarkdown')
m <- data.frame('Value\n1', 'Value\n2')
colnames(m) <- c('Multiline\nCol1', 'Multiline\nCol2')
pandoc.table(m, keep.line.breaks = TRUE, style='rmarkdown')
```

## Cell alighnment

## Highlighting cells

## Table and Cell width

## Minor options

Hyphening, plain.ascii, caption and etc.

## panderOptions



