---
title: "Randering markdown with pander"
author: "Roman Tsegelskyi, Gergely Daróczi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pandoc.table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(pander)
library(tables)
panderOptions('knitr.auto.asis', FALSE)
panderOptions('plain.ascii', TRUE)
```

This vignette introduces `pander` package and it's core pieces of functionality. It is intented to be a general overview with pointers to where to get more detailed information. 

## Pander

`Pander` is R package designed to provide a minimal and easy tool for rendering `R` objects into [Pandoc](http://pandoc.org/)'s markdown. `pander` is also capable of exporting/converting complex `Pandoc` documents (reports) in various ways.

## Rendering R objects

As mentioned above core functionality of `pander` is centered around rendering `R` objects into `Pandoc`'s markdown. Let's dive in and demo the most common usage of `pander`:

```{r}
pander(head(iris))
pander(head(mtcars[1:5]))
pander(tabular( (Species + 1) ~ (n=1) + Format(digits=2)*
         (Sepal.Length + Sepal.Width)*(mean + sd), data=iris ))
```

This is achieved via generic `pander` `S3` method. Out of the box, `pander` supports a variety of classes:

```{r}
methods(pander)
```


If you think that pander lacks support for any other R class(es), please feel free to leave change request here and we will be happy to extend the package.

Under the hood, `pander` relies on different `pandoc.*` methods, where most of functionality is implemented in `pandoc.table` which is used for rendering tables. `pandoc.table` provides a truly rich functionality with many different options that `pander` inherits. For more detail please refer to specialized vignetted, which can be accessed by `vignette('pandoc_table')` or available online [here]().

### Pander and knitr

## Evals

## General Options
