---
title: "Randering markdown with pander"
author: "Roman Tsegelskyi, Gergely Daróczi"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pandoc.table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(collapse = T, comment = "#>")
library(pander)
library(tables)
panderOptions('knitr.auto.asis', FALSE)
panderOptions('plain.ascii', TRUE)
```

`Pander` is designed to provide a minimal and easy tool for rendering `R` objects into [Pandoc](http://pandoc.org/)'s markdown. This vignette aims to introduce `pander` package and it's core pieces of functionality. It is intented to be a general overview with pointers to places with detailed information. This vignette will talk about:

* core functionality for rendering objects in `Pandoc`'s markdown with generic S3 [pander method](#rendering-r-objects).
* functionality for capturing various information when evaluating R expressions with [evals](#evals).
* report generation with [Pandoc.brew](#brew-to-pandoc).
* globally adjustable options through [panderOptions/evalsOptions](#general-options).

## Rendering R objects

The core functionality of `pander` is centered around rendering `R` objects into `Pandoc`'s markdown. Let's dive in the demo of the most common usage of `pander`:

```{r}
pander(head(iris))
pander(head(mtcars[1:5]))
pander(tabular( (Species + 1) ~ (n=1) + Format(digits=2)*
         (Sepal.Length + Sepal.Width)*(mean + sd), data=iris ))
```

As you have probably guesse, this is achieved via generic `pander` `S3` method. Out of the box, `pander` supports a variety of classes:

```{r}
methods(pander)
```

If you think that pander lacks support for any other R class(es), please feel free to leave change request [here](https://github.com/Rapporter/pander/issues) and we will be happy to extend the package.

Under the hood, `pander` S3 methods rely on different `pandoc.*` methods, where most of functionality is implemented in `pandoc.table` which is used for rendering tables. `pandoc.table` provides  a truly rich functionality with many different options that `pander` inherits. For more usage/implementation details and examples, please refer to specialized vignette, which can be accessed by `vignette('pandoc_table')` or available online [here]().

## Evals

As `pander` package was originally developed in conjunction with [rapport](https://github.com/Rapporter/rapport) package, there was a needed for functionality that can evaluate `R` expression along with capturing errors and warnings. So `evals` emerged and soon some further feature requests arose, like identifying if an R expression results in a plot etc.

But probably it's easier to explain what `evals` can do with a simple example:

```{r}
evals('1:10')
```

`evals` is aimed at collecting as much information as possible while evaluating R code. It can evaluate a character vector of R expressions, and it returns a list of captured information while running those:

* `src` holds the R expression,
* `result` contains the raw R object as is,
* `output` represents how the R object is printed to the standard output,
* `type` is the class of the returned R object,
* `msg` is a list of possible messages captured while evaluating the R expression and. Among other messages, warnings/errors will appear here.
* `stdout` contains if anything was written to the standard output.

For more usage/implementation details and examples, please refer to specialized vignette, which can be accessed by `vignette('evals')` or available online [here]().

## Brew to Pandoc

The [brew](https://cran.r-project.org/web/packages/brew/index.html) package, which is a templating framework for report generation, has not been updated since 2011, but it's still some of R projects based on its simple design and useful features in literate programming. For a quick overview, please see the following documents if you are not familiar with brew:

* [slides on "Building a reporting sytem with BREW"](http://www.slideshare.net/xavierguardiola/building-a-reporting-sytem-with-brew)
* [learnr blogpost on brew](https://learnr.wordpress.com/2009/09/09/brew-creating-repetitive-reports/)

A brew document is a simple text file with some special tags. `Pandoc.brew` uses only two of them (as building on a personalized version of Jeff's really great brew function):

* `<% ... %>` stand for running inline R commands as usual,
* `<%= ... %>` does pretty much the same but applies pander to the returning R object (instead of cat like the original brew function does). So putting there any R object, it would return in a nice Pandoc's markdown format with all possible error/warning messages etc.

This latter tries to be smart in some ways:

* A code chunk block (`R` commands between the tags) can return any number of values at any part of the block.
* Plots and images are grabbed in the document, rendered to a png file and pander method would result in a Pandoc markdown formatted image link. This means that the image would be rendered/shown/included in the exported document.
* All warnings/messages and errors are recorded in the blocks and returned in the document as footnotes or inline messages.
* All heavy `R` commands (e.g. those taking more then 0.1 sec to evaluate) are cached so rebrewing a report would not result in a coffee break.

Besides this, the custom brew function can do more and also less compared to the original brew package. First of all, the internal caching mechanism of brew has been removed and rewritten for some extra profits besides improved caching. Quick example:

```{r}
str(Pandoc.brew(text ='
+   Pi equals to `<%= pi %>`.
+   And here are some random data:
+   `<%= runif(10) %>`
+ '))
```

The package bundles some examples for `Pandoc.brew` to let you check its features pretty fast. 
To brew these examples on your machine, try to run the followings commands:

```
Pandoc.brew(system.file('examples/minimal.brew', package='pander'))
Pandoc.brew(system.file('examples/minimal.brew', package='pander'),
            output = tempfile(), convert = 'html')

Pandoc.brew(system.file('examples/short-code-long-report.brew', package='pander'))
Pandoc.brew(system.file('examples/short-code-long-report.brew', package='pander'),
                        output = tempfile(), convert = 'html')

Pandoc.brew(system.file('examples/graphs.brew', package='pander'))
Pandoc.brew(system.file('examples/graphs.brew', package='pander'),
                        output = tempfile(), convert = 'html')
```

## General Options

The package comes with a variety of globally adjustable options, which have an effect on the result of your reports. You can query and update these options with the `panderOptions` function:

  * `digits`: numeric (default: `2`) passed to `format`
  * `decimal.mark`: string (default: `.`) passed to `format`
  * `formula.caption.prefix`: string (default: `Formula: `) passed to `pandoc.formula` to be used as caption prefix. Be sure about what you are doing if changing to other than `Formula: ` or `:`.
  * `big.mark`: string (default: ``) passed to `format`
  * `round`: numeric (default: `Inf`) passed to `round`
  * `keep.trailing.zeros`: boolean (default: `FALSE`) show or remove trailing zeros in numbers (e.g. in numeric vectors or in columns of tables with numeric values)
  * `keep.line.breaks`: boolean (default: `FALSE`) to keep or remove line breaks from cells in a table
  * `missing`: string (default: `NA`) to replace missing values in vectors, tables etc.
  * `date`: string (default: `'%Y/%m/%d %X'`) passed to `format` when printing dates (`POSIXct` or `POSIXt`)
  * `header.style`: `'atx'` or `'setext'` passed to `pandoc.header`
  * `list.style`: `'bullet'` (default), `'ordered'` or `'roman'` passed to `pandoc.list`. Please not that this has no effect on `pander` methods.
  * `table.style`: `'multiline'`, `'grid'` or `'simple'` passed to `pandoc.table`
  * `table.emphasize.rownames`: boolean (default: `TRUE`) if row names should be highlighted
  * `table.split.table`: numeric passed to `pandoc.table` and also affects `pander` methods. This option tells `pander` where to split too wide tables. The default value (`80`) suggests the conventional number of characters used in a line, feel free to change (e.g. to `Inf` to disable this feature) if you are not using a VT100 terminal any more :)
  * `table.split.cells`: numeric (default: 30) passed to `pandoc.table` and also affects pander methods. This option tells pander where to split too wide cells with line breaks. Set `Inf`` to disable.
  * `table.caption.prefix`: string (default: `Table: `) passed to `pandoc.table` to be used as caption prefix. Be sure about what you are doing if changing to other than `Table: ` or `:`.
  * `table.continues`: string (default: `Table continues below`) passed to `pandoc.table` to be used as caption for long (split) without a use defined caption
  * `table.continues.affix`: string (default: `(continued below)`) passed to `pandoc.table` to be used as an affix concatenated to the user defined caption for long (split) tables
  * `table.alignment.default`: string (default: `centre`) that defines the default alignment of cells. Can be `left`, `right` or `centre` that latter can be also spelled as `center`
  * `table.alignment.rownames`: string (default: `centre`) that defines the alignment of rownames in tables. Can be `left`, `right` or `centre` that latter can be also spelled as `center`
  * `use.hyphening`: boolean (default: `FALSE`) if try to use hyphening when splitting large cells according to table.split.cells. Requires `koRpus` package.
  * `evals.messages`: boolean (default: `TRUE`) passed to `evals`' `pander` method specifying if messages should be rendered
  * `p.wrap`: a string (default:`'_'`) to wrap vector elements passed to `p` function
  * `p.sep`: a string (default: `', '`) with the main separator passed to `p` function
  * `p.copula`: a string (default: `'and'`) a string with ending separator passed to `p` function
  * `plain.ascii`: boolean (default: FALSE) to define if output should be in plain ascii or not
  * `graph.nomargin`: boolean (default: `TRUE`) if trying to keep plots' margins at minimal
  * `graph.fontfamily`: string (default: `'sans'`) specifying the font family to be used in images. Please note, that using a custom font on Windows requires `grDevices:::windowsFonts` first.
  * `graph.fontcolor`: string (default: `'black'`) specifying the default font color
  * `graph.fontsize`: numeric (default: `12`) specifying the *base* font size in pixels. Main title is rendered with `1.2` and labels with `0.8` multiplier.
  * `graph.grid`: boolean (default: `TRUE`) if a grid should be added to the plot
  * `graph.grid.minor`: boolean (default: `TRUE`) if a miner grid should be also rendered
  * `graph.grid.color`: string (default: `'grey'`) specifying the color of the rendered grid
  * `graph.grid.lty`: string (default: `'dashed'`) specifying the line type of grid
  * `graph.boxes`: boolean (default: `FALSE`) if to render a border around of plot (and e.g. around strip)
  * `graph.legend.position`: string (default: `'right'`) specifying the position of the legend: 'top', 'right', 'bottom' or 'left'
  * `graph.background`: string (default: `'white'`) specifying the plots main background's color
  * `graph.panel.background`: string (default: `'transparent'`) specifying the plot's main panel background. Please *note*, that this option is not supported with `base` graphics.
  * `graph.colors`: character vector of default color palette (defaults to a [colorblind theme](http://jfly.iam.u-tokyo.ac.jp/color/)). Please *note* that this update work with `base` plots by appending the `col` argument to the call if not set.
  * `graph.color.rnd`: boolean (default: `FALSE`) specifying if the palette should be reordered randomly before rendering each plot to get colorful images
  * `graph.axis.angle`: numeric (default: `1`) specifying the angle of axes' labels. The available options are based on `par(les)` and sets if the labels should be:

      *  `1`: parallel to the axis,
      *  `2`: horizontal,
      *  `3`: perpendicular to the axis or
      *  `4`: vertical.

  * `graph.symbol`: numeric (default: `1`) specifying a symbol (see the `pch` parameter of `par`)
  * `knitr.auto.asis`: boolean (default: `TRUE`) if the results of `pander` should be considered as `asis` in `knitr`. Equals to specifying `results='asis'` in the R chunk, so thus there is no need to do so if set tot `TRUE`.

<a id='evals-options'></a><a id='evalsoptions'></a>

Besides localization of numeric formats or the styles of tables, lists and plots, there are some technical options as well, which would effect e.g. [caching](#caching) or the format of rendered image files. You can query/update those with the `evalsOptions` function as the main backend of `pander` calls is a custom evaluation function called [`evals`](#evals).

The list of possible options are:

  * `parse`: if `TRUE` the provided `txt` elements would be merged into one string and parsed to logical chunks. This is useful if you would want to get separate results of your code parts - not just the last returned value, but you are passing the whole script in one string. To manually lock lines to each other (e.g. calling a `plot` and on next line adding an `abline` or `text` to it), use a plus char (`+`) at the beginning of each line which should be evaluated with the previous one(s). If set to `FALSE`, [`evals`](#evals) would not try to parse R code, it would get evaluated in separate runs - as provided. Please see the documentation of [`evals`](#evals).
  * `cache`: [caching](#caching) the result of R calls if set to `TRUE`
  * `cache.mode`: cached results could be stored in an `environment` in _current_ R session or let it be permanent on `disk`.
  * `cache.dir`: path to a directory holding cache files if `cache.mode` set to `disk`. Default set to `.cache` in current working directory.
  * `cache.time`: number of seconds to limit caching based on `proc.time`. If set to `0`, all R commands, if set to `Inf`, none is cached (despite the `cache` parameter).
  * `cache.copy.images`: copy images to new file names if an image is returned from the *disk* cache? If set to `FALSE` (default), the cached path would be returned.
  * `classes`: a vector or list of classes which should be returned. If set to `NULL` (by default) all R objects will be returned.
  * `hooks`: list of hooks to be run for given classes in the form of `list(class = fn)`. If you would also specify some parameters of the function, a list should be provided in the form of `list(fn, param1, param2=NULL)` etc. So the hooks would become `list(class1=list(fn, param1, param2=NULL), ...)`. See example of [`evals`](#evals) for more details. A default hook can be specified too by setting the class to `'default'`. This can be handy if you do not want to define separate methods/functions to each possible class, but automatically apply the default hook to all classes not mentioned in the list. You may also specify only one element in the list like: `hooks=list('default' = pander.return)`. Please note, that nor error/warning messages, nor stdout is captured (so: updated) while running hooks!
  * `length`: any R object exceeding the specified length will not be returned. The default value (`Inf`) does not filter out any R objects.
  * `output`: a character vector of required returned values. This might be useful if you are only interested in the `result`, and do not want to save/see e.g. `messages` or `print`ed `output`. See examples of [`evals`](#evals).
  * `graph.unify`: should `evals` try to unify the style of (`base`, `lattice` and `ggplot2`) plots? If set to `TRUE`, some `panderOptions()` would apply. By default this is disabled not to freak out useRs :)
  * `graph.name`: set the file name of saved plots which is `%s` by default. A simple character string might be provided where `%d` would be replaced by the index of the generating `txt` source, `%n` with an incremented integer in `graph.dir` with similar file names and `%t` by some unique random characters. When used in a `brew` file, `%i` is also available which would be replaced by the chunk number.
  * `graph.dir`: path to a directory where to place generated images. If the directory does not exist, [`evals`](#evals) try to create that. Default set to `plots` in current working directory.
  * `graph.output`: set the required file format of saved plots. Currently it could be any of  `grDevices`: `png`, `bmp`, `jpeg`, `jpg`, `tiff`, `svg` or `pdf`. Set to `NA` not to save plots at all and tweak that setting with `capture.plot()` on demand.
  * `width`: width of generated plot in pixels for even vector formats
  * `height`: height of generated plot in pixels for even vector formats
  * `res`: nominal resolution in `ppi`. The height and width of vector images will be calculated based in this.
  * `hi.res`: generate high resolution plots also? If set to `TRUE`, each R code parts resulting an image would be run twice.
  * `hi.res.width`: width of generated high resolution plot in pixels for even vector formats. The `height` and `res` of high resolution image is automatically computed based on the above options to preserve original plot aspect ratio.
  * `graph.env`: save the environments in which plots were generated to distinct files (based on `graph.name`) with `env` extension?
  * `graph.recordplot`: save the plot via `recordPlot` to distinct files (based on `graph.name`) with `recodplot` extension?
  * `graph.RDS` save the raw R object returned (usually with `lattice` or `ggplot2`) while generating the plots to distinct files (based on `graph.name`) with `RDS` extension?
  * `log`: `NULL` or  an optionally passed *logger name* from `futile.logger` to record all info, trace, debug and error messages.
